### 2022-03-08 scratch.rmd
early musings

```{r}
library(gssr)
library(tidyverse)
# don't recommend getting single year -- not enough data
# gss21 <- gss_get_yr(2021)
# gss21["sexsex"]
data(gss_all)

```
```{r}
gss_clean <- gss_all |>
  mutate(
    id = paste0(year, '_', id),
    sexnow = coalesce(sexnow, sexnow1),
    sexbirth = coalesce(sexbirth, sexbirth1)
    ) |>
  select(-sexnow1, -sexbirth1)
```

```{r}
# grab a subset of data, specifically
# all responses from years in which all
# required questions in {qns} are asked/
# have answers.
# Optionally, also filter to only entries
# with answers to all questions.
getSubsetWithAnswers <- function(qns, clean_na = FALSE) {
  years_all_asked <- gss_clean |>
    gss_which_years(qns) |>
    na_if(FALSE) |>
    drop_na() |>
    pull(year)

  gss_clean |>
    filter(year %in% years_all_asked) |>
    select(year, id, all_of(qns)) |>
    filter(across(qns, ~ !clean_na | !is.na(.)))
}
```

```{r}
# this dataset contains a whopping 1 intersex person,
# and only 12 who identify their SEXNOW as 3 [transgender].
# In fact, only 42 people identify their SEXBIRTH as different from SEXNOW.
getSubsetWithAnswers(c("sexnow", "sexbirth"), clean_na = T) |>
  ggplot(aes(x = as.factor(sexbirth), y = as.factor(sexnow))) +
  geom_bin2d()
```
```{r}
# Interesting...
gss_clean |> 
  filter(sexbirth != sexnow |
         sexnow != sex |
         sex != sexbirth) |> 
  select(id, sexbirth, sexnow, sex)
```

### 2022-03-14 scratch.rmd
added faceted graph with a few ideas

@@
 # gss21["sexsex"]
-data(gss_all)
 
 ```
+
+```{r}
+data(gss_all)
+```
+
 ```{r}
@@
     id = paste0(year, '_', id),
-    sexnow = coalesce(sexnow, sexnow1),
-    sexbirth = coalesce(sexbirth, sexbirth1)
+    sexnow = coalesce(sexnow, sexnow1) |> as_factor(),
+    sexbirth = coalesce(sexbirth, sexbirth1) |> as_factor(),
+    sexornt = as_factor(sexornt),
+    sexsex = as_factor(sexsex),
+    sexsex5 = as_factor(sexsex5)
     ) |>
-  select(-sexnow1, -sexbirth1)
+  select(id, sexnow, sexbirth, sexornt, sexsex, sexsex5)
+```
+```{r}
+
 ```
@@
 ```
+
+```{r}
+gss_clean |>
+  filter(!if_any(c(sexsex, sexnow, sexornt), is.na)) |>
+  ggplot(aes(x = sexsex, fill=sexsex)) +
+  geom_histogram(stat="count") +
+  scale_fill_manual(values=c('blue', 'purple', 'red')) +
+  theme(axis.text.x = element_blank()) +
+  facet_grid(sexnow ~ sexornt,
+             scales = "free_y")
+```
+```{r}
+gss_clean |>
+  filter(!if_any(c(sexsex5, sexnow, sexornt), is.na)) |>
+  ggplot(aes(x = sexsex5, fill=sexsex5)) +
+  geom_histogram(stat="count") +
+  scale_fill_manual(values=c('blue', 'purple', 'red')) +
+  #geom_text(stat="count", aes(label=..count..), position = position_stack(vjust=.5), color="black") +
+  theme_bw() + 
+  theme(
+    axis.text.x = element_blank(),
+    axis.ticks.x = element_blank(),
+    panel.grid.major.x = element_blank()
+    ) +
+  facet_grid(sexornt ~ sexnow,
+             scales = "free_y",
+             labeller = label_wrap_gen(width = 16, multi_line = T)
+             )
+```

### 2022-03-25 scratch.rmd
Add more exploratory graphs

--- 2022-03-08 scratch.rmd
@@
     sexnow = coalesce(sexnow, sexnow1),
-    sexbirth = coalesce(sexbirth, sexbirth1)
+    sexbirth = coalesce(sexbirth, sexbirth1),
     ) |>
-  select(-sexnow1, -sexbirth1)
+  select(id, year, sex, sexnow, sexbirth, sexornt, sexsex, sexsex5)
+```
+```{r}
+
 ```
@@
 getSubsetWithAnswers <- function(qns, clean_na = FALSE) {
-  years_all_asked <- gss_clean |>
+  years_all_asked <- gss_all |>
     gss_which_years(qns) |>
@@
 
-  gss_clean |>
+  gss_all |>
     filter(year %in% years_all_asked) |>
@@
 ```
+
+```{r}
+gss_clean |>
+  filter(!if_any(c(sexsex, sexnow, sexornt), is.na)) |>
+  ggplot(aes(x = as_factor(sexsex), fill=as_factor(sexsex))) +
+  geom_histogram(stat="count") +
+  scale_fill_manual(values=c('blue', 'purple', 'red')) +
+  theme(axis.text.x = element_blank()) +
+  facet_grid(sexnow ~ sexornt,
+             scales = "free_y")
+```
+```{r}
+gss_clean |>
+  filter(!if_any(c(sexsex5, sexnow, sexornt), is.na)) |>
+  ggplot(aes(x = as_factor(sexsex5), fill=as_factor(sexsex5))) +
+  geom_histogram(stat="count") +
+  scale_fill_manual(values=c('blue', 'purple', 'red')) +
+  #geom_text(stat="count", aes(label=..count..), position = position_stack(vjust=.5), color="black") +
+  theme_bw() + 
+  theme(
+    axis.text.x = element_blank(),
+    axis.ticks.x = element_blank(),
+    panel.grid.major.x = element_blank()
+    ) +
+  facet_grid(sexornt ~ sexnow,
+             scales = "free_y",
+             labeller = label_wrap_gen(width = 16, multi_line = T)
+             )
+```
+
+```{r}
+# Well, let's see just /how/ flawed the data is.
+
+subsetWith <- function(..) {
+  gss_clean |>
+    filter(!if_any(.., is.na))
+}
+```
+```{r}
+gender_counts <- subsetWith(c('sex', 'sexbirth', 'sexnow')) |>
+  mutate(
+    sex_changed = sexbirth != sexnow,
+    sex_mismatch = sexbirth == sexnow & sexnow != sex,
+    sex_coded_birth = sex == sexbirth,
+    sex_coded_now = sex == sexnow
+  ) |>
+  count(sex, sexbirth, sexnow, sex_changed, sex_mismatch, sex_coded_birth, sex_coded_now) |>
+  mutate(
+  trans_miscoded = sex_changed & !sex_coded_now
+  )
+
+gender_counts |>
+  ggplot(aes(x = as_factor(trans_miscoded), y = n, fill=trans_miscoded)) +
+  geom_bar(stat = "identity") +
+  labs(
+    x = "Respondent-identified gender",
+    y = "Interviewer-coded sex of respondent",
+    fill = "Apparent misgendering?"
+  ) +
+   theme(
+    axis.ticks.x = element_blank(),
+    axis.text.x = element_blank()
+  ) +
+  facet_grid(as_factor(sex) ~ as_factor(sexnow), scales="free_y", labeller = label_wrap_gen(width = 16, multi_line = T))
+
+# gss_clean |> 
+#   filter(!if_any(c(sex, sexbirth, sexnow), is.na)) |>
+#   mutate(
+#     sex_mismatch = 
+#   )
+#   group_by(year) |>
+#   summarize()
+#   ggplot(aes(x = factor(year)))
+```
+```{r}
+gender_counts |>
+  #filter(sexnow %in% c(1, 2)) |>
+  ggplot(aes(x = sex_coded_now, y = n, fill = sex_coded_now)) +
+  geom_bar(stat = "identity", position="dodge") +
+  geom_text(aes(label=n, vjust="center")) +
+  labs(
+    x = "Assigned sex at birth (respondent-identified)",
+    y = "Gender identity (respondent-identified)*",
+    fill = "Reporter-indicated sex matches gender?"
+  ) +
+   theme(
+    axis.ticks.x = element_blank(),
+    axis.text.x = element_blank()
+  ) +
+  #scale_fill_manual(values=c('blue', 'red')) + 
+  facet_grid(as_factor(sexnow) ~ as_factor(sexbirth), scales="free_y", labeller = label_wrap_gen(width = 16, multi_line = T))
+```

### 2022-03-26 paper_3.Rmd
beginning of work mar 26

---
title: "Self-identification of gender and sexual orientation in the US General Social Survey:"
subtitle: "A review of methodological inconsistencies and anomalous data"
abstract: >
  The US General Social Survey, administered by the National Opinion Research Center at the University of Chicago, was recently emended to include options for respondents to specify several aspects of their sexual orientation and gender identity, including how their gender has changed relative to their birth sex. However, upon exploration of the data, it became clear that several major inconsistencies and anomalies existed in both the methodology by which the data were collected, as well as in the data themselves. This paper highlights these issues, and offers as an appendix a supplementary survey that would help to rectify these anomalies in future General Social Surveys.
author: "Oliver Daniel & Tiago Martins"
date: "due March 20, 2022[^1]"
geometry: "margin=1.5in"
knit: (function(input, ...) {
    rmarkdown::render(
    input,
    output_file = "out/paper_3.pdf",
    envir = globalenv(),
    ...
    )
  })
bibliography: in/references.bib
link-citations: true
---

```{r setup, echo=FALSE}
knitr::opts_chunk$set(echo = FALSE, out.width = "80%", out.align="center", dir)
options(knitr.graphics.error = FALSE)
```

```{r libraries, include=FALSE}
if (!require(tidyverse)) {install.packages('tidyverse')}
if (!require(gssr)) {remotes::install_github('https://github.com/kjhealy/gssr')}

library(tidyverse)
library(gssr)
```

```{r data-setup}
data(gss_all)
gss_clean <- gss_all |>
  mutate(
    id = paste0(year, '_', id),
    sexnow = coalesce(sexnow, sexnow1),
    sexbirth = coalesce(sexbirth, sexbirth1),
    sexsex = sexsex,
    sexsex5 = sexsex5,
    ornt_sex_label = paste(
      case_when(
        sexornt == 1 ~ "Gay",
        sexornt == 2 ~ "Bisexual",
        sexornt == 3 ~ "Straight"
      ),
      case_when(
        sexnow == 1 ~ "man",
        sexnow == 2 ~ "woman",
        sexnow == 3 ~ "trans person",
        sexnow == 4 ~ "person of another gender",
      )
    ),
    .keep = "used"
  )
```

```{r functions}
subsetWith <- function(..., sel=F) {
  result <- gss_clean |>
    filter(!if_any(c(...), is.na))

  if (sel) result <- result |>
      select(...)
  result
}
```

## Introduction
TODO
## Data
### Dataset
This paper focuses on the United States General Social Survey (US-GSS, or GSS in the the following), a survey regularly administered by the National Opinion Research Center (NORC) at the University of Chicago [@norc-gss]. For reasons elaborated below, although the cumulative data file contains `r nrow(gss_clean) |> format(big.mark=',')` observations across `r gss_clean$year |> n_distinct()` years of study, we will be focusing primarily on only respondents from the last two surveys, in 2018 and 2021. This significantly limits certain aspects of our investigation to `r gss_clean |> filter(year >= 2018) |> nrow() |> format(big.mark =',')` entries; however, the issues we highlight are endemic to the format and methodology of the survey itself, and thus do not require large magnitudes to afford predictive power.

In most years, the GSS is conducted in the form of a face-to-face interview, in which an interviewer poses a particular subset of questions in series to a respondent, and encodes their answer into a regularized format. To this end, the respondent is usually prompted with several options, e.g., "Agree", "Neither agree nor disagree", "Disagree", from which they choose the most accurate response. In addition, some variables allow for 'volunteered' responses, which are codeable responses that are *not* read to the respondent, but rather only recorded if the respondent independently volunteers said response.

However, due to the constraints of the COVID-19 pandemic, the 2020 survey – which was postponed into 2021 – was partially conducted through an online survey platform (and the remainder via telephone). This platform did not offer volunteered responses, and respondents could only choose from the set of options presented to them. For 2021, the GSS bifurcated these variables in two, using either a -V or -NV suffix to indicate whether the respondent's particular mode of communication afforded volunteered responses. As it happens, none of these variables are of study in this paper; however, as we are examining something so sensitive to wording as gender, it stands to reason that the absence of volunteered responses could have been detrimental to the granularity of description offered to respondents with which to communicate their identity. For example, the variable `SEXNOW1` – the 1 suffix indicating this specific phrasing of the question was *only* asked during the pandemic – only offers four options: "Male", "Female", "Transgender", or "None of these". This does not form a complementary set: someone identifying, say, as a transgender woman would be required to respond either "Female" or "Transgender", but not both. If "transgender woman" had been available as a volunteered response, it is possible that some of the issues caused by this encoding would have been avoided.

The official GSS Codebook [@gss-codebook] indicates that the variables of the survey consist of a combination of respondent-offered responses (i.e., respondents' answers to a given question) and interviewer-coded responses (i.e., questions that are not directly asked of the respondent, but rather recorded *about* the respondent, at the interviewer's discretion). Again, for something as personal as gender or sexual identity, this necessarily introduces a cognitive gap that is subject to the interviewer's personal perceptions and biases. As a pathological example, a particularly malicious interviewer could knowingly and willingly mis-code a transgender respondent's sex (under the `SEX` variable) as their birth sex, rather than their chosen gender. As a result, we chose to de-emphasize `SEX` — an interviewer-coded variable asked across every ballot of every GSS since 1972 – in our study, in favour of the respondent-focused `SEXNOW` and `SEXNOW1`. These variables are only asked in a single year each – 2018 and 2021, respectively – which *significantly* reduces our pool of respondents. Future studies may be interested in a longitudinal view of how these new, more flexible questions of identity shape responses over time.

### Variables
Although 2021-specific variables like `SEXNOW1` are necessarily only answered by a small fraction of respondents, they use the same coding as their pre-2021 equivalents, and exist in complementary distribution to them (i.e., no respondent has a response to both variables). As a result, we were able to expand our selection by coalescing the variables together into a single column. This is notated below with a `(1)` suffix, although the 1-suffixed column was dropped in the actual dataframe. With this in mind, our summative list of variables is as follows:

- `YEAR`: The survey year in which the respondent was interviewed.
- `SEX`: The **interviewer-coded** sex of the respondent. (Used only for statistics on reporting, not for respondent identity.) Options: 1 (Male), 2 (Female).
- `SEXBIRTH(1)`: The assigned sex of the respondent at birth, based on physiognomy. Options: 1 (Male), 2 (Female), 3 (Intersex).
- `SEXNOW(1)`: The personally-identified sex/gender of the respondent, at survey time. Options: 1 (Male), 2 (Female), 3 (Transgender), 4 (None of these).
- `SEXORNT`: The sexual[^3] orientation of the respondent. Options: 1 (Gay/Lesbian/Homosexual), 2 (Bisexual), 3 (Straight/Heterosexual).
- `SEXSEX`: The gender makeup of the respondents' sexual partners, to their knowledge, for the past 12 months. Options: 1 (Exclusively male), 2 (Both male and female), 3 (Exclusively female).
- `SEXSEX5`: Similar to `SEXSEX`, but expanded over the past 5 years. Options: 1 (Exclusively male), 2 (Both male and female), 3 (Exclusively female).

### Computed Variables
In addition to the above data variables, several additional columns were computed to make the associated data more portable and readable at-a-glance.

- `id`: A serial ID. Not used directly, but it was discovered that the serial IDs for respondents rolledback to 1 every survey year. For reproducibility, it was important to ensure that there were no ID collisions. So, the serial ID is now also prefixed with the year.
- `sex_ornt_label`: A compact, human-readable description of a person's gender and sexual orientation, as might be used in everyday conversation. e.g., "Straight woman", "Gay man", "Bisexual trans person".[^4]

### Methods
All data analysis[^2] was performed in the R statistical programming language [@R], using RStudio [@RStudio] for ease of exploration, development, and production of this paper via built-in `knitr` [@knitr] functionality. GSS data are fetched by the `gssr` package [@gssr], which automatically packages the columns of the data into a ready-to-use dataframe. As usual, the `tidyverse` [@tidyverse] collection provides an indispensable array of resources, including but not limited to `ggplot2` [@ggplot2] for visualization.

## Results
TODO
## Discussion
TODO
## Survey
TODO

\newpage
## References

<!-- Footnotes -->
[^1]: Extended.
[^2]: Code available at https://github.com/oliver-daniel/inf313_paper_3.
[^3]: There is an important distinction to be made between sexual attraction and romantic attraction, not to mention sexual attraction valency (i.e., the asexuality spectrum). However, for the purposes of this study and supplementary survey, only the prior is considered.
[^4]: Note that "gay" is used to refer to any kind of exclusively-homosexual attraction, as identified by the respondent. Historically, "gay" often referred exclusively to male homosexual attraction, with "lesbian" being more common for females; conversely, "gay" is currently entering common parlance for *any* same-gender attraction, even among those who also experience different-gender attraction. The usage of terms such as these to refer to different intersections of gender and sexuality over time is fascinating, and worthy of its own study; however, it is immaterial to this study, and so "gay" was chosen as a cover-all term.

### 2022-03-26 paper_3.Rmd
Lots more graphs

@@
 ```{r setup, echo=FALSE}
-knitr::opts_chunk$set(echo = FALSE, out.width = "80%", out.align="center", dir)
+knitr::opts_chunk$set(echo = FALSE, warning = FALSE, out.width = "80%", out.align="center", dir)
 options(knitr.graphics.error = FALSE)
@@
     id = paste0(year, '_', id),
+    sex = sex,
     sexnow = coalesce(sexnow, sexnow1),
@@
     sexsex5 = sexsex5,
-    ornt_sex_label = paste(
+    ptnrornt = ptnrornt,
+    ornt_sex_label = if_else(
+      is.na(sexornt) | is.na(sexnow),
+      as.character(NA),
+      paste(
       case_when(
@@
       )
-    ),
+    )),
     .keep = "used"
-  )
+  ) |>
+  select(-one_of(c('sexnow1', 'sexbirth1')))
 ```
@@
 TODO
+
 ## Data
@@
 
+Finally, an important note on the usage of `NA` in the dataset: although the Codebook includes marginals for reserved non-response codes – "Don't know", "No answer", "Skipped on web" – the software used to compile the data into an R dataframe [@gssr] condense all of these into the singular value of `NA`.
+
 ### Variables
@@
 - `SEXORNT`: The sexual[^3] orientation of the respondent. Options: 1 (Gay/Lesbian/Homosexual), 2 (Bisexual), 3 (Straight/Heterosexual).
-- `SEXSEX`: The gender makeup of the respondents' sexual partners, to their knowledge, for the past 12 months. Options: 1 (Exclusively male), 2 (Both male and female), 3 (Exclusively female).
+- `SEXSEX`: The gender makeup of the respondent's sexual partners, to their knowledge, for the past 12 months. Options: 1 (Exclusively male), 2 (Both male and female), 3 (Exclusively female).
 - `SEXSEX5`: Similar to `SEXSEX`, but expanded over the past 5 years. Options: 1 (Exclusively male), 2 (Both male and female), 3 (Exclusively female).
+- `PTNRORNT`: The orientation of the respondent's *latest* sexual partner, to their knowledge.
+Options: 1 (Gay/Lesbian/Homosexual), 2 (Bisexual), 3 (Straight/Heterosexual), 4 (I have never had a sexual partner)[^4].
 
@@
 - `sex_ornt_label`: A compact, human-readable description of a person's gender and sexual orientation, as might be used in everyday conversation.
-e.g., "Straight woman", "Gay man", "Bisexual trans person".[^4]
+e.g., "Straight woman", "Gay man", "Bisexual trans person".[^5]
 
@@
 As usual, the `tidyverse` [@tidyverse] collection provides an indispensable array of resources, including but not limited to `ggplot2` [@ggplot2] for visualization.
+For additional aesthetics, `kableExtra` [@kableExtra], which styles tables, and `patchwork` [@patchwork], which composes multiple `ggplot` plots together, were also used.
 
 ## Results
-TODO
+
+```{r fig-demographics}
+subsetWith('year', 'sexnow', 'sexornt', sel = T) |>
+  ggplot(aes(x = as_factor(sexnow), y = after_stat(count), fill = as_factor(sexornt))) +
+  geom_bar() +
+  facet_wrap(~year) +
+  theme(
+    axis.text.x = element_text(angle=20, vjust=.95, hjust=1),
+    legend.position = "bottom"
+  ) +
+  labs(
+    title = "Gender and sexuality identification demographics of GSS respondents across 2018 and 2021 surveys" |> str_wrap(60),
+    x = "Gender identity of respondent",
+    y = "Count",
+    fill = "Sexual orientation of respondent"
+  )
+```
+```{r fig-demographics-trans-nb}
+subsetWith('year', 'sexnow', 'sexornt', sel = T) |>
+  filter(sexnow > 2) |>
+  ggplot(aes(x = as_factor(sexnow), y = after_stat(count), fill = as_factor(sexornt))) +
+  geom_bar() +
+  facet_wrap(~year) +
+  theme(
+    axis.text.x = element_text(angle=20, vjust=.95, hjust=1),
+    legend.position = "bottom"
+  ) +
+  labs(
+    title = "Demographics of transgender and non-binary GSS respondents across 2018 and 2021 surveys" |> str_wrap(60),
+    x = "Gender identity of respondent",
+    y = "Count",
+    fill = "Sexual orientation of respondent"
+  )
+```
+
+```{r table-birth-now}
+
+counts <- 
+  subsetWith('sexbirth', 'sexnow', sel=T) |> 
+  table()
+
+rownames(counts) <- c('Assigned male', 'Assigned female', 'Intersex')
+
+knitr::kable(
+    counts,
+    col.names = c('Male', 'Female', 'Transgender', 'Other'),
+  ) |>
+  kableExtra::kable_classic(full_width = FALSE)
+```
+
+```{r table-sex-coding}
+birth_codes <- c('M', 'F', 'I')
+gender_codes <- c('M', 'F', 'T', 'X')
+
+subsetWith('sexbirth', 'sexnow') |>
+  mutate(
+    code = paste0(birth_codes[sexbirth], gender_codes[sexnow]),
+    ) |>
+  select(code, sex) |>
+  table() |>
+  knitr::kable(col.names = c('Recorded as male', 'Recorded as female')) |>
+  kableExtra::kable_classic(full_width = FALSE)
+```
+
+```{r fig-sexual-partners}
+library(patchwork)
+
+makePlot <- function(y_var) {
+  subsetWith('ornt_sex_label', all_of(y_var), sel=T) |>
+    ggplot(aes(y = reorder(ornt_sex_label, ornt_sex_label, length), fill = as_factor(.data[[all_of(y_var)]]))) +
+    geom_bar(position = position_fill(reverse=TRUE))
+}
+
+makePlot('ptnrornt') + labs(title = '"My latest sexual partner was..."') + theme(legend.position = "bottom", legend.direction = "vertical", axis.title.y = element_blank(), axis.title.x = element_blank(), legend.title = element_blank(), axis.ticks.y = element_blank())
+```
+```{r}
+# ((makePlot('sexsex') + theme(axis.title.y = element_blank())) +
+# (makePlot('sexsex5') + theme(axis.text.y = element_blank(), axis.title.y = element_blank()))) +
+(makePlot('sexsex') + labs(title = "Over the past 12 months")) +
+(makePlot('sexsex5') + labs(title = "Over the past 5 years") + theme(axis.text.y = element_blank())) +
+  plot_layout(guides = "collect") &
+  theme(legend.position = "bottom", axis.title.x = element_blank(), axis.title.y = element_blank(), legend.title = element_blank(), axis.ticks.y = element_blank())
+```
+
 ## Discussion
 TODO
+
 ## Survey
@@
 However, for the purposes of this study and supplementary survey, only the prior is considered.
-[^4]: Note that "gay" is used to refer to any kind of exclusively-homosexual attraction, as identified by the respondent.
+[^4]: This is the only variable under investigation which includes an option for "I have never had a sexual partner".
+As a result, most analyses involving this variable will explicitly filter out this option, to keep a level playing field among the other sex-partner-related questions.
+[^5]: Note that "gay" is used to refer to any kind of exclusively-homosexual attraction, as identified by the respondent.
 Historically, "gay" often referred exclusively to male homosexual attraction, with "lesbian" being more common for females; conversely, "gay" is currently entering common parlance for *any* same-gender attraction, even among those who also experience different-gender attraction.

### 2022-03-27 paper_3.Rmd
near end of work mar 26

--- 2022-03-26 paper_3.Rmd
@@
 geometry: "margin=1.5in"
+output:
+  bookdown::pdf_document2:
+    toc: true
+    number_sections: false
+    fig_crop: false
+    extra_dependencies: ["flafter"]
 knit: (function(input, ...) {
@@
 
+```{r preamble}
+#### Preamble ####
+# Purpose: Fetch, clean, and present data from the NORC United States General Social Survey, knitted into a research paper
+# Authors: Oliver Daniel and Tiago Martins
+# Date: 2022-03-20
+# Contact: via Quercus or email
+# License: WTFPL
+# Prerequisites:
+# - None known; running this file should auto-install
+# dependencies.
+
+#TODO: move setup to another script?
+```
+
 ```{r setup, echo=FALSE}
-knitr::opts_chunk$set(echo = FALSE, out.width = "80%", out.align="center", dir)
+knitr::opts_chunk$set(echo = FALSE, warning = FALSE, out.width = "80%", out.align="center", dir)
 options(knitr.graphics.error = FALSE)
@@
 if (!require(gssr)) {remotes::install_github('https://github.com/kjhealy/gssr')}
+if (!require(kableExtra)) {install.packages('kableExtra')}
+if (!require(patchwork)) {install.packages('patchwork')}
 
 library(tidyverse)
+library(kableExtra)
+library(patchwork)
 library(gssr)
@@
       case_when(
-        sexnow == 1 ~ "man",
-        sexnow == 2 ~ "woman",
-        sexnow == 3 ~ "trans person",
-        sexnow == 4 ~ "person of another gender",
+        sexnow == 1 ~ "men",
+        sexnow == 2 ~ "women",
+        sexnow == 3 ~ "trans people",
+        sexnow == 4 ~ "people of other genders",
       )
-    ),
+    )),
     .keep = "used"
-  )
+  ) |>
+  select(-one_of(c('sexnow1', 'sexbirth1')))
 ```
@@
 
+```{r colours}
+colour <- list(
+  male = '#0273f5',
+  intersex = '#9B02ED',
+  female = '#fb0036',
+  trans = '#F495D1',
+  other = '#098000',
+  gay = '#A50314',
+  bisexual = '#B759D9',
+  straight = '#0375A6',
+  none = '#464646'
+)
+
+```
+
 ## Introduction
 TODO
+
 ## Data
@@
 As usual, the `tidyverse` [@tidyverse] collection provides an indispensable array of resources, including but not limited to `ggplot2` [@ggplot2] for visualization.
+For additional aesthetics, `kableExtra` [@kableExtra], which[^6] styles tables, and `patchwork` [@patchwork], which composes multiple `ggplot` plots together, were also used.
 
 ## Results
-TODO
+
+```{r fig-demographics, fig.cap="\\label{fig:demographics} In both 2018 and 2021, respondent demographics are dominated by straight men and women.
+A significantly larger contingent of bisexual women responded in 2021."}
+subsetWith('year', 'sexnow', 'sexornt', sel = T) |>
+  ggplot(aes(x = as_factor(sexnow), y = after_stat(count), fill = as_factor(sexornt))) +
+  geom_bar() +
+  facet_wrap(~year) +
+  theme(
+    axis.text.x = element_text(angle=20, vjust=.95, hjust=1),
+    axis.title.y = element_blank(),
+    legend.position = "bottom",
+    legend.direction = "vertical"
+  ) +
+  labs(
+    title = "Gender and sexuality identification demographics of GSS respondents across 2018 and 2021 surveys" |> str_wrap(60),
+    x = "Gender identity of respondent",
+    fill = "Sexual orientation of respondent"
+  ) +
+  scale_fill_manual(values=c(colour$gay, colour$bisexual, colour$straight))
+```
+```{r fig-demographics-trans-nb, fig.cap="\\label{fig:demographics-trans-nb} The overall turnout of transgender and 'other' (e.g., non-binary) respondents is low, but exhibits a wider diversity in sexuality."}
+subsetWith('year', 'sexnow', 'sexornt', sel = T) |>
+  filter(sexnow > 2) |>
+  ggplot(aes(x = as_factor(sexnow), y = after_stat(count), fill = as_factor(sexornt))) +
+  geom_bar() +
+  facet_wrap(~year) +
+  theme(
+    axis.text.x = element_text(angle=20, vjust=.95, hjust=1),
+    axis.title.y = element_blank(),
+    legend.position = "bottom"
+  ) +
+  labs(
+    title = "Demographics of transgender and 'other' GSS respondents across 2018 and 2021 surveys" |> str_wrap(60),
+    x = "Gender identity of respondent",
+    fill = "Sexual orientation of respondent"
+  ) +
+  scale_fill_manual(values=c(colour$gay, colour$bisexual, colour$straight))
+```
+
+```{r table-birth-now}
+
+counts <- 
+  subsetWith('sexbirth', 'sexnow', sel=T) |> 
+  table()
+
+rownames(counts) <- c('Assigned male', 'Assigned female', 'Intersex')
+
+knitr::kable(
+    counts,
+    col.names = c('Male', 'Female', 'Transgender', 'Other'),
+    caption = "As expected, cisgendered respondents vastly outnumber any other configuration; in this dataset, the entire intersex population of the United States is represented by a single respondent."
+  )
+```
+More than 22% of self-reported cisgender women are recorded as male, and 30% of self-reported cisgender men as female.
+```{r table-sex-coding}
+birth_codes <- c('M', 'F', 'I')
+gender_codes <- c('M', 'F', 'T', 'X')
+
+counts <- subsetWith('sexbirth', 'sexnow') |>
+  mutate(
+    code = paste0(birth_codes[sexbirth], '/', gender_codes[sexnow]),
+    ) |>
+  select(code, sex) |>
+  table()
+
+colnames(counts) <- c('Recorded as male', 'Recorded as female')
+
+  knitr::kable(
+    counts,
+    caption="This table exhibits the first major inconsistency in the data."
+    )
+```
+
+```{r fig-latest-sexual-partners, fig.cap="\\label{fig:latest-sexual-partners}Surprisingly, the distributions of this variable seem to be more accurately reflective than those in Figure \\ref{fig:sexual-partners-over-time}, in terms of what sorts of sexual partners one would expect a respondent to have had."}
+makePlot <- function(y_var) {
+  subsetWith('ornt_sex_label', all_of(y_var), sel=T) |>
+    ggplot(aes(y = reorder(ornt_sex_label, ornt_sex_label, length), fill = as_factor(.data[[all_of(y_var)]]))) +
+    geom_bar(position = position_fill(reverse=TRUE))
+}
+
+makePlot('ptnrornt') + theme(legend.position = "bottom", legend.direction = "vertical", axis.title.y = element_blank(), axis.title.x = element_blank(), legend.title = element_blank(), legend.justification = "left", axis.ticks.y = element_blank()) + plot_annotation(
+  title = '"In terms of their orientation, my latest sexual partner was..."',
+  subtitle = "by personal identity, as a proportion of respondents"
+) + scale_fill_manual(values=c(colour$gay, colour$bisexual, colour$straight, colour$none))
+```
+
+```{r fig-sexual-partners-over-time, fig.cap="\\label{fig:sexual-partners-over-time}Although preference in sexual partners appears relatively stable over both 1-year and 5-year periods, an anomalously large contingent of heterosexual-identifying men and women have had exclusively same-sex relationships over these periods."}
+(makePlot('sexsex') + labs(title = "over the past 12 months")) +
+(makePlot('sexsex5') + labs(title = "over the past 5 years") + theme(axis.text.y = element_blank())) +
+  plot_layout(guides = "collect")  +
+  plot_annotation(title = '"In terms of their sex, my sexual partners have been..."',
+  subtitle = "by personal identity, as a proportion of respondents") &
+  theme(legend.position = "bottom", axis.title.x = element_blank(), axis.title.y = element_blank(), legend.title = element_blank(), axis.ticks.y = element_blank()) &
+  scale_fill_manual(values=c(colour$male, colour$bisexual, colour$female))
+```
+
 ## Discussion
 TODO
+
 ## Survey
@@
 However, for the purposes of this study and supplementary survey, only the prior is considered.
-[^4]: Note that "gay" is used to refer to any kind of exclusively-homosexual attraction, as identified by the respondent.
+[^4]: This is the only variable under investigation which includes an option for "I have never had a sexual partner".
+As a result, most analyses involving this variable will explicitly filter out this option, to keep a level playing field among the other sex-partner-related questions.
+[^5]: Note that "gay" is used to refer to any kind of exclusively-homosexual attraction, as identified by the respondent.
 Historically, "gay" often referred exclusively to male homosexual attraction, with "lesbian" being more common for females; conversely, "gay" is currently entering common parlance for *any* same-gender attraction, even among those who also experience different-gender attraction.
 The usage of terms such as these to refer to different intersections of gender and sexuality over time is fascinating, and worthy of its own study; however, it is immaterial to this study, and so "gay" was chosen as a cover-all term.
+[^6]: Attempts to, anyway.
+Half of the options cause Kable to dump raw \LaTeX  and destroy the formatting of the entire document.
+Did you know that it's literally impossible to move table captions to the bottom of the table?

### 2022-03-27 paper_3.Rmd
check in work mar 27

@@
 
-```{r preamble}
+```{r preamble, echo = FALSE}
 #### Preamble ####
@@
 if (!require(gssr)) {remotes::install_github('https://github.com/kjhealy/gssr')}
+if (!require(knitr)) {remotes::install_github('yihui/knitr')}
 if (!require(kableExtra)) {install.packages('kableExtra')}
 if (!require(patchwork)) {install.packages('patchwork')}
+if (!require(ggalluvial)) {install.packages('ggalluvial')}
 
 library(tidyverse)
+library(ggalluvial)
 library(kableExtra)
@@
 data(gss_all)
+
+lvls <- expand.grid(x = c("Gay", "Bisexual", "Straight"), y = c("men", "women", "trans people", "people of other genders")) |> arrange(y)
+ordered_levels <- do.call(paste, lvls)
+
 gss_clean <- gss_all |>
@@
     ptnrornt = ptnrornt,
+    ptnrsxnow = ptnrsxnow,
     ornt_sex_label = if_else(
@@
       paste(
-      case_when(
-        sexornt == 1 ~ "Gay",
-        sexornt == 2 ~ "Bisexual",
-        sexornt == 3 ~ "Straight"
-      ),
-      case_when(
-        sexnow == 1 ~ "men",
-        sexnow == 2 ~ "women",
-        sexnow == 3 ~ "trans people",
-        sexnow == 4 ~ "people of other genders",
-      )
-    )),
+        levels(lvls[,1])[sexornt],
+        levels(lvls[,2])[sexnow]
+    )) |> factor(levels = ordered_levels),
+    ptnr_label = if_else(
+      is.na(ptnrornt) | is.na(ptnrsxnow),
+      as.character(NA),
+      paste(
+        levels(lvls[,1])[ptnrornt],
+        levels(lvls[,2])[ptnrsxnow]
+      )) |> factor(levels = ordered_levels),
     .keep = "used"
@@
 Options: 1 (Gay/Lesbian/Homosexual), 2 (Bisexual), 3 (Straight/Heterosexual), 4 (I have never had a sexual partner)[^4].
+- `PTNRSXNOW`: The gender identity of the respondent's *latest* sexual partner, to their knowledge. Options: 1 (Male), 2 (Female), 3 (Transgender), 4 (None of these).
 
@@
 So, the serial ID is now also prefixed with the year.
-- `sex_ornt_label`: A compact, human-readable description of a person's gender and sexual orientation, as might be used in everyday conversation.
+- `sex_ornt_label`: A compact, human-readable description of a respondent's gender and sexual orientation, as might be used in everyday conversation.
 e.g., "Straight woman", "Gay man", "Bisexual trans person".[^5]
+- `ptnr_label`: Similar, for respondents' latest sexual partner.
 
@@
 ## Results
+
+### Respondent demographics
+We first sought to develop an understanding of the demography of our respondents.
+In Figure \@ref(fig:demographics) below, we see that respondents to the 2018 and 2021 GSS overwhelmingly self-identify as men and women (as opposed to transgendered or another gender), and predominantly as heterosexual.
+Interestingly, although the bisexual-identifying segment of the 2018 female respondent pool is so small as to be almost invisible, a significantly greater magnitude and proportion of them are present in the 2021 pool.
+The same trend is visible in the population of gay men, albeit to a lesser extent.
 
@@
 ```
+Figure \@ref(fig:demographics-trans-nb) zooms in on the non-binary gender options which, although representing an extremely small portion of the population – with no cohort representing more than ten respondents in a given year – exhibits a greater diversity in sexuality than their binary counterparts.
+
 ```{r fig-demographics-trans-nb, fig.cap="\\label{fig:demographics-trans-nb} The overall turnout of transgender and 'other' (e.g., non-binary) respondents is low, but exhibits a wider diversity in sexuality."}
@@
 ```
+With access to not only respondents' present gender identity, but also their birth sex, we were interested to determine the distinction, if any, between those respondents who responded 'transgender' as opposed to a binary gender.
+Table \@ref(tab:table-birth-now) contrasts birth sex (rows) with gender identity (columns) and counts total respondents at each intersection.
+As expected, the majority of respondents identify as cisgendered: that is, that their present gender corresponds to their birth sex.
+As many male-assigned respondents identified as female (i.e., as trans women) as transgender, implying that there is not a clear semantic delineation between the two.
+And, with the presence of a fourth 'other' category – which includes two male-assigned respondents – it is not necessarily the case that 'transgender' also includes non-binary and gender-non-conforming identities.
+This linguistic distinction is a clear area for future study.
+Fascinatingly, despite making up an estimated [TODO: amount, cite] of the U.S.
+population, only one respondent intimated that they were born intersex.
+As a result, and lacking access to this individual's actual interview/questionnaire, we will focus less on the complicated realm of post-intersex gender identity, especially as it corresponds to post-natal gender reassignment by medical practitioners; this, too, is a viable subject of intense study.
+[TODO CITE?]
 
@@
 ```
-More than 22% of self-reported cisgender women are recorded as male, and 30% of self-reported cisgender men as female.
+
+### Interviewer gender coding and the `SEX` variable
+
+As mentioned previously, early explorations of the data necessitated disregarding the `SEX` variable almost entirely.
+This is perhaps most strikingly exhibited below in Table \@ref(tab:table-sex-coding); for brevity, the first column delineates the birth sex (M = male; F = female; I = intersex) and the gender identification (M = male; F = female; T = transgender; X = other) with a slash.
+
 ```{r table-sex-coding}
@@
 ```
+As shown, more than 22% of self-reported cisgender women are recorded by interviewers as male, and 30% of self-reported cisgender men as female.
+Without further access to the specific methodologies of conducting a US-GSS interview, it is unknown to us what could have caused such a significant anomaly.
+The Codebook [@gss-codebook] notes:
+
+> The variable SEX was revised for the web mode as well. In face-to-face interviews, interviewers traditionally coded
+SEX based on interviewer observation. In the web mode, SEX was asked explicitly of respondents. Sex and gender
+identity are collected as two separate items: sex recorded at birth (SEXBIRTH1) and current gender identity
+(SEXNOW1). In the past, these items have only been asked to two-thirds of respondents (i.e., two ballots), on a
+self-administered topical module. Beginning with the 2020/2021 GSS (both the 2016-2020 GSS Panel Wave 2 and
+the 2021 GSS Cross-section), SEXBIRTH1 and SEXNOW1 are asked of all respondents. For the purpose of
+backward compatibility, the cross-section dataset contains SEX, which uses the traditional binary coding scheme,
+and is based on recoding both SEXBIRTH1 and SEXNOW1 in 2021 (which was collected by interviewer observation
+in prior years).
+
+It is not at all clear what is meant by 'recoding', in terms of how the `SEX` column is calculated from the aforementioned respondent-identified columns, especially when this calculation produced more than one-fifths' error in what should have been the two most straightforward cases.
+It furthermore reflects poorly on the NORC that in each non-cisgendered combination, at least half of respondents are coded as their birth sex, even when their chosen gender is explicitly different.
+For example, 100% of female-assigned respondents who identified as an unlisted gender were recorded as female, as were 50% of male-assigned, female respondents.
+Although we hesitate to draw from this a malicious reading, this inconsistency in gender-recording procedure, across multiple intersections of sex and gender, is certainly cause for suspicion.
+
+### Sexual orientation and recent sex partners
+Tables \@ref(fig:partner-correlation)-\@ref(fig:sexual-partners-over-time) investigate the gender and sexuality makeup of respondents' recent sexual partners.
+The spans in question range from only the latest engagement, to a five-year period.
+
+Figure \@ref(fig:partner-correlation) below provides a bird's-eye view of respondents' most recent sexual activity.
+Although a single sexual partner does not necessarily exemplify *all* of a respondent's partners over time, the recency of the data hopefully mitigates the risk of one's historical sexual behaviour being irreflective of their present orientation, especially in a Western heteronormative environment.
+This means, for example, that a person may have had several partners of the opposite sex before realizing that they were gay or otherwise interested in the same sex.
+As a result, their latest partner is likely – though by no means certain – to be someone to whose gender and sexuality the respondent is attracted.
+
+```{r fig-partner-correlation, fig.cap="\\label{fig:partner-correlation}Including a log scale was necessary to compete with the overwhelmingly frequent pairings of straight men and straight women."}
+subsetWith('ornt_sex_label', 'ptnr_label', sel = T) |>
+  select(ornt_sex_label, ptnr_label) |>
+  ggplot(aes(x = ptnr_label, y = ornt_sex_label, fill = log(after_stat(count)))) +
+  scale_x_discrete(limits=ordered_levels) +
+  geom_bin2d() +
+  coord_fixed() +
+  scale_fill_viridis_c(option="C") +
+  labs(
+    title = str_wrap("Log-scale correlation heatmap of respondent's identity to that of their latest sexual partner", width = 60),
+    y = "Identity of respondent",
+    x = '"To my knowledge, my last sexual partner was..."',
+    fill = "frequency\n(log scale)"
+  ) +
+  theme(axis.text.x = element_text(angle = 45,hjust=1))
+```
+Many high-frequency pairings are expected: straight men with straight women, gay men with gay men, and so on.
+Bisexual woman respondents in particular exhibit a wider variety in their partners than any other cohort of respondents, including some in which the partners' fields of attraction appear to mismatch.
+In one case, for example, a respondent who identified as a bisexual woman responded that their latest partner was a straight woman.
+In real life, such sexual pairings absolutely happen [TODO CITE?
+maybe some queer theory?] Similarly, a number of sexual encounters were recorded between straight men and straight men (`r gss_clean |> filter(ornt_sex_label == "Straight men" & ptnr_label == "Straight men") |> nrow()`), and straight women with straight women (`r gss_clean |> filter(ornt_sex_label == "Straight wo" & ptnr_label == "Straight women") |> nrow()`).
+Surprisingly, transgender and non-binary (i.e., 'other') respondents exclusively reported binary-gendered partners.
+However, it must be acknowledged that, as such a narrow slice of the respondent pool, this absence could be by sheer chance.
+
+Narrowing our focus to the sexuality (or existence) of the respondent's latest partner appears to confirm this: although binary-gendered respondents show at least some diversity in the proportions of their partners' sexualities, each row in Figure \@ref(fig:latest-sexual-partners) below corresponding to a transgender or nonbinary cohort exhibits either 100% or 50% consensus, indicating very low magnitudes.
 
@@
   subsetWith('ornt_sex_label', all_of(y_var), sel=T) |>
-    ggplot(aes(y = reorder(ornt_sex_label, ornt_sex_label, length), fill = as_factor(.data[[all_of(y_var)]]))) +
+    ggplot(aes(y = ornt_sex_label, fill = as_factor(.data[[all_of(y_var)]]))) +
     geom_bar(position = position_fill(reverse=TRUE))
@@
 ```
+However, we run into another surprising anomaly when analyzing variables `SEXSEX` and `SEXSEX5`, which concern longitudinal sexual behaviour over 1- and 5-year periods, respectively.
+As might be expected, the 5-year results are usually slightly more diverse than their shorter-term counterparts; bisexual respondents, in particular, responded 'Both male and female' more frequently over the expanded timeframe.
+The sexual partners of straight men, on the other hand, appear hardly to change at all, save for a slightly-widened sliver of 'Both male and female' responses.
+That being said, one would imagine that these partners would be exclusively women – or that this would be the case for most straight men.
+Instead, almost 30% of straight male respondents report having had sex exclusively with men.
+Similarly, nearly a third of straight women report having sex only with women in the past year, a proportion that increases to full half when considering the past 5 years.
+Again, although we do not suppose to claim such a result is *impossible*, a 22-year-long observational study into the behaviour of straight people (Daniel et al, 2022)[^7] appears to indicate this is at the least an *improbable* result.
+Instead, we propose that this is another notational error, similar to the miscoding of cisgendered respondents discussed above.
 
@@
   scale_fill_manual(values=c(colour$male, colour$bisexual, colour$female))
+```
+
+```{r}
+subsetWith('sex', 'sexnow', 'sexornt', 'sexsex', 'ornt_sex_label', sel=T) |>
+  filter(sex <= 2 & sexornt != 2) |>
+  mutate(
+    sex_miscoded = sex != sexnow,
+    unexpected_partner = case_when(
+      sex == 1 & sexornt == 3 ~ sexsex != 3,
+      sex == 1 & sexornt == 1 ~ sexsex != 1,
+      sex == 2 & sexornt == 3 ~ sexsex != 1,
+      sex == 2 & sexornt == 1 ~ sexsex != 1,
+      TRUE ~ as.logical(NA)
+    )
+  ) |> 
+  #filter(sex_miscoded | unexpected_par)
+  group_by(ornt_sex_label, sex_miscoded, unexpected_partner) |>
+  count()
 ```
@@
 Did you know that it's literally impossible to move table captions to the bottom of the table?
+[^7]: Not cited; paper does not exist.

### 2022-03-28 paper_3.Rmd
Continued work mar 27

@@
 if (!require(patchwork)) {install.packages('patchwork')}
-if (!require(ggalluvial)) {install.packages('ggalluvial')}
 
 library(tidyverse)
-library(ggalluvial)
 library(kableExtra)
@@
 In real life, such sexual pairings absolutely happen [TODO CITE?
-maybe some queer theory?] Similarly, a number of sexual encounters were recorded between straight men and straight men (`r gss_clean |> filter(ornt_sex_label == "Straight men" & ptnr_label == "Straight men") |> nrow()`), and straight women with straight women (`r gss_clean |> filter(ornt_sex_label == "Straight wo" & ptnr_label == "Straight women") |> nrow()`).
+maybe some queer theory?] Similarly, a number of sexual encounters were recorded between straight men and straight men (`r gss_clean |> filter(ornt_sex_label == "Straight men" & ptnr_label == "Straight men") |> nrow()`), and straight women with straight women (`r gss_clean |> filter(ornt_sex_label == "Straight women" & ptnr_label == "Straight women") |> nrow()`).
 Surprisingly, transgender and non-binary (i.e., 'other') respondents exclusively reported binary-gendered partners.
@@
 ```
+Table \@ref(tab:table-miscodings-unexpected) investigates this anomaly further, specifically as it pertains to binary-gendered respondents with a single-sex attraction, e.g., respondents who identify as male or female, and gay or straight.
+In summary, it appears that having one's gender misrecorded in `SEX` but having had sexual partners exclusively of the expected sex, or vice versa, is significantly less likely than both variables being either expected or unexpected in value.
+To illustrate this confusing wording with an example, there are significantly more self-identified gay men who:
 
-```{r}
-subsetWith('sex', 'sexnow', 'sexornt', 'sexsex', 'ornt_sex_label', sel=T) |>
-  filter(sex <= 2 & sexornt != 2) |>
+- a) were recorded as *men* and b) reported having sex with *only men* in the past year; or
+- a) were recorded as *women* and b) reported having sex with *only women* in the past year
+
+than some combination of the two.
+Mysteriously, gay women – the least abundant of the cohorts – reverse this trend, being the only cohort with any respondents who were misrecorded as men, but only having reported sex with women, and zero respondents to be accurately recorded as women with only female partners.
+This appears to be evidence of a comparatively widespread scrambling of responses when compared to personal identification, which may well have interfered with other findings in this paper.
+It is unknown at this time whether this confusion stems from survey respondents themselves, the interviewers who recorded their responses, at some point in the NORC datafiles, or even in the `gssr` client (@gssr) we are using to fetch and process the data.
+
+```{r table-miscodings-unexpected}
+
+subsetWith('sex', 'sexnow', 'sexornt', 'ornt_sex_label', 'sexsex', sel=F) |>
+filter(sexnow <= 2 & sexornt %in% c(1, 3)) |>
   mutate(
@@
     unexpected_partner = case_when(
-      sex == 1 & sexornt == 3 ~ sexsex != 3,
-      sex == 1 & sexornt == 1 ~ sexsex != 1,
-      sex == 2 & sexornt == 3 ~ sexsex != 1,
-      sex == 2 & sexornt == 1 ~ sexsex != 1,
-      TRUE ~ as.logical(NA)
+      ornt_sex_label == "Straight men" ~ sexsex != 3,
+      ornt_sex_label == "Gay men" ~ sexsex != 1,
+      ornt_sex_label == "Straight women" ~ sexsex != 1,
+      ornt_sex_label == "Gay women" ~ sexsex != 1,
+      TRUE ~ FALSE
     )
   ) |>
-  #filter(sex_miscoded | unexpected_par)
-  group_by(ornt_sex_label, sex_miscoded, unexpected_partner) |>
-  count()
+  count(ornt_sex_label, sex_miscoded, unexpected_partner) |>
+  pivot_wider(names_from = c(sex_miscoded, unexpected_partner), values_from = n, values_fill = 0) |>
+  relocate(TRUE_TRUE, .after = FALSE_TRUE) |>
+  kable(col.names = c('Identity', 'No/No', 'No/Yes', 'Yes/Yes', 'Yes/No'),
+        caption = "There appears to be a correlative effect between a mismatch in recorded sex, and the likelihood of sexual experiences beyond the expected demographic.
+Apart from that, there seems to be a confounding effect afoot."
+        ) |>
+  kable_classic() |>
+  add_header_above(header = c(" " = 1, "Sex miscoded?/Unexpected partner?" = 4))
 ```
@@
 
-## Survey
-TODO
+## Appendix: Survey
+Our survey can be found online [here](https://forms.gle/AsAQ8nudjoSHc6sz9).
 

### 2022-03-28 scripts/scratch.rmd
Reshuffle folders + README

--- 2022-03-25 scratch.rmd
@@
     sexbirth = coalesce(sexbirth, sexbirth1),
-    ) |>
-  select(id, year, sex, sexnow, sexbirth, sexornt, sexsex, sexsex5)
+    )# |>
+  #select(id, year, sex, sexnow, sexbirth, sexornt, sexsex, sexsex5, ptnrornt)
 ```
@@
 ```{r}
-gss_clean |>
-  filter(!if_any(c(sexsex, sexnow, sexornt), is.na)) |>
+subsetWith(c('sexsex', 'sexnow', 'sexornt')) |>
   ggplot(aes(x = as_factor(sexsex), fill=as_factor(sexsex))) +
@@
   theme(axis.text.x = element_blank()) +
-  facet_grid(sexnow ~ sexornt,
-             scales = "free_y")
+  facet_grid(as_factor(sexnow) ~ as_factor(sexornt),
+             scales = "free_y",
+             labeller = label_wrap_gen(width = 16, multi_line = T)
+             )
 ```
@@
     ) +
-  facet_grid(sexornt ~ sexnow,
+  facet_grid(as_factor(sexornt) ~ as_factor(sexnow),
              scales = "free_y",
@@
 ```
+
+```{r}
+sex_codes <- c('M', 'F', 'I')
+gender_codes <- c('M', 'F', 'T', 'X')
+
+subsetWith(c('sex', 'sexbirth', 'sexnow')) |>
+  mutate(
+    code = paste0(sex_codes[sexbirth], gender_codes[sexnow])
+    ) |>
+  select(code, sex) |>
+  group_by(code) |>
+  table() |>
+  knitr::kable(col.names = c("Coded male", "Coded female")) |>
+  kableExtra::kable_classic()
+```
+
+```{r}
+subsetWith(c('sexornt', 'ptnrornt')) |>
+  select('sexornt', 'ptnrornt') |>
+  filter(ptnrornt != 4) |> # 'I've never had a sexual partner'
+  group_by(sexornt) |>
+  ggplot(aes(x = as_factor(sexornt), y = as_factor(ptnrornt))) + 
+  geom_bin_2d() +
+  scale_fill_viridis_c(option="C")
+```
+
+```{r}
+subsetWith(c('sexnow', 'sexornt', 'ptnrornt', 'ptnrsxnow')) |>
+  select('sexnow', 'sexornt', 'ptnrornt', 'ptnrsxnow') |>
+  ggplot(aes(x = as_factor(ptnrornt), fill = as_factor(ptnrsxnow))) +
+  geom_histogram(stat='count') +
+  theme_bw() +
+  theme(axis.text.x = element_blank()) +
+  labs(
+    title = "Gender identity of subject",
+    x = "Sexual orientation of last sex partner (Gay, Bisexual, Straight)",
+    y = "Sexual orientation of subject",
+    fill = "Gender identity of last sex partner"
+  ) +
+  facet_grid(as_factor(sexornt) ~ as_factor(sexnow), 
+  scale="free_y",          
+  labeller = label_wrap_gen(width = 16, multi_line = T))
+```
+
+```{r}
+# based on `sex`, how many respondents have primarily been having
+# sexual relationships with an unexpected cohort? (e.g., straight men primarily having sex with men)
+subsetWith('sex', 'sexornt', 'sexsex') |>
+  select(sex, sexornt, sexsex) |>
+  mutate(
+    expected_sexsex = case_when(
+      sex == 1 & sexornt == 3 ~ 3,
+      sex == 1 & sexornt == 1 ~ 1,
+      sex == 2 & sexornt == 3 ~ 1,
+      sex == 2 & sexornt == 1 ~ 3,
+    )
+  ) |>
+  filter(
+    #!is.na(expected_sexsex) &
+    expected_sexsex != sexsex
+  )
+```
+```{r}
+# sim. with sexnow and sexsex5
+subsetWith('sexnow', 'sexornt', 'sexsex5') |>
+  select(year, sexnow, sexornt, sexsex5) |>
+  rowwise() |>
+  mutate(
+    label = paste(
+      c("Gay", "Bisexual", "Straight")[sexornt],
+      c("men", "women", "transgender people", "people of other genders")[sexnow],
+      "primarily having sex with",
+      c("men", "women", "both men and women")[sexsex5]
+    , collapse = " "),
+    expected_sexsex = case_when(
+      sexnow == 1 & sexornt == 3 ~ 3,
+      sexnow == 1 & sexornt == 1 ~ 1,
+      sexnow == 2 & sexornt == 3 ~ 1,
+      sexnow == 2 & sexornt == 1 ~ 3,
+    )
+  ) |>
+  count(label)
+
+# |>
+#   filter(
+#     #!is.na(expected_sexsex) &
+#     sexsex != expected_sexsex
+#   )
+```

### 2022-03-28 paper_3.Rmd
Update README
(hopefully) final commit

@@
 # dependencies.
-
-#TODO: move setup to another script?
 ```
@@
 ```{r setup, echo=FALSE}
-knitr::opts_chunk$set(echo = FALSE, warning = FALSE, out.width = "80%", out.align="center", dir)
+knitr::opts_chunk$set(echo = FALSE, warning = FALSE, out.width = "100%", out.align="center", dir)
 options(knitr.graphics.error = FALSE)
@@
 ## Introduction
-TODO
+The General Social Survey, also known as the General Social Surveys, is a series of interviews conducted by the National Opinion Research Center at the University of Chicago [@gssr].
+These surveys are cross-sectional, meaning that they analyze data across a population at a specific point in time.
+The General Social Survey, henceforth referred to as GSS, covers a variety of subject matter, including topics relating to sex, gender identity, and sexual partners.
+It is these variables that are the subject matter of this paper.
+
+Utilizing the programming language R [@R], the integrated development environment RStudio [@RStudio], and several associated packages, we have drawn from the 2019 GSS data set and constructed several graphs to visualize trends in the areas of sex and gender.
+In doing so, we found several issues with the structuring of survey questions, the most egregious being a lack of gender expression and identity response options, particularly with regards to transgender individuals.
+We discuss the effects this has, while also noting the effects of the COVID-19 pandemic on this iteration of the GSS.
+
+Lastly, as part of our analysis of the shortcomings in the 2019 GSS, we provide examples of similar data collection methods that offered potential solutions to these issues.
+While they are from two different fields, New Zealand’s Statistical Standard for Gender Identity [@Northland], and the Centers for Disease Control and Prevention’s Behavioural Risk Factor Surveillance System [@Healthcare], both offer more nuanced options for interviewee gender identity responses.
+In the appdendix, we include a survey of our creation which, if adopted by the NORC, would be of benefit to future GSS studies.
 
@@
 This linguistic distinction is a clear area for future study.
-Fascinatingly, despite making up an estimated [TODO: amount, cite] of the U.S.
-population, only one respondent intimated that they were born intersex.
-As a result, and lacking access to this individual's actual interview/questionnaire, we will focus less on the complicated realm of post-intersex gender identity, especially as it corresponds to post-natal gender reassignment by medical practitioners; this, too, is a viable subject of intense study.
-[TODO CITE?]
+Likewise, only one respondent intimated that they were born intersex; as a result, and lacking access to this individual's actual interview/questionnaire, we will focus less on the complicated realm of post-intersex gender identity, especially as it corresponds to post-natal gender reassignment by medical practitioners; this, as @Northland outlines, is a viable subject of intense study.
 
@@
 In one case, for example, a respondent who identified as a bisexual woman responded that their latest partner was a straight woman.
-In real life, such sexual pairings absolutely happen [TODO CITE?
-maybe some queer theory?] Similarly, a number of sexual encounters were recorded between straight men and straight men (`r gss_clean |> filter(ornt_sex_label == "Straight men" & ptnr_label == "Straight men") |> nrow()`), and straight women with straight women (`r gss_clean |> filter(ornt_sex_label == "Straight women" & ptnr_label == "Straight women") |> nrow()`).
+In real life, such sexual pairings absolutely happen [@NotGay], but are less common than these data would indicate.
+Similarly, a number of sexual encounters were recorded between straight men and straight men (`r gss_clean |> filter(ornt_sex_label == "Straight men" & ptnr_label == "Straight men") |> nrow()`), and straight women with straight women (`r gss_clean |> filter(ornt_sex_label == "Straight women" & ptnr_label == "Straight women") |> nrow()`).
 Surprisingly, transgender and non-binary (i.e., 'other') respondents exclusively reported binary-gendered partners.
@@
 This appears to be evidence of a comparatively widespread scrambling of responses when compared to personal identification, which may well have interfered with other findings in this paper.
-It is unknown at this time whether this confusion stems from survey respondents themselves, the interviewers who recorded their responses, at some point in the NORC datafiles, or even in the `gssr` client (@gssr) we are using to fetch and process the data.
+It is unknown at this time whether this confusion stems from survey respondents themselves, the interviewers who recorded their responses, at some point in the NORC datafiles, or even in the `gssr` client [@gssr] we are using to fetch and process the data.
 
@@
 ## Discussion
-TODO
+While unlikely that this was the intention of the GSS creators, one cannot help but see effects of inadequate response options in a survey intended to be nationally representative.
+As previously mentioned, the options for gender self-identification in the 2019 GSS comprised of merely four categories: male, female, transgender, and other [@norc-gss].
+As there exists within the data individuals of particular birth sexes identifying as members of the opposite sex, as part of their gender identity, one questions the applicability of the transgender identity option at all.
+And as always, there is the ever-present risk that the transferring from a live interview to a self-administered digital survey may have greatly altered participant responses due to the shift in medium.
+
+It should be known that, in the collection of data on gender and sexuality, the articulation of the meaning of categories is important.
+Sullivan [@Conflate] gives the example of the misconception of using the terms ‘sex’ and ‘gender’ synonymously, thereby crippling the sources of data used for the studying of various social issues.
+One such issue is the existence of son preference within certain countries, and with it, the risk of sex-targeted abortions.
+Without the knowledge of an individual’s birth sex, how are we continue to study and combat issues that target specific birth sexes?
+Sullivan’s work offers a critique of the conflation between these terms, something the GSS has thus far avoided.
+
+However, in a continuation of the discussion regarding gender identity options, we can look to other examples of surveys which offer different perspectives on the collection of this data.
+The Government of New Zealand’s Statistical Standard for Gender Identity, an overhauled version of the previous Statistical Standard for Sex, outlines the parameters for which data regarding gender identity is to be gathered.
+This statistical standard seeks to facilitate the standardization of term definitions and the measurements of gender identity [@Northland] so that continually developing human rights requirements are met.
+The Standard defines gender identity as: “an individual’s internal sense of being wholly female, wholly male, or having aspects of female and male” [@Northland].
+Its identity options are also novel compared to the GSS, offering the options of: male, female, and gender diverse, which has the participant fill in the most appropriate terminology for themselves.
+While there are indeed criticisms of this approach, the fact that it is this legislature that is the first ever national statistical standard for gender identity, shows that there are still new avenues open for other nations to be more representative.
+
+Another example of a different approach to gathering information on gender identities is the Behavioural Risk Factor Surveillance System, abbreviated as BRFSS, which are cross-sectional, population-based health surveys overseen by the Centers for Disease Control and Prevention [@Healthcare].
+The BRFSS as a whole is composed of several modules, one of them being dedicated to gender identity and sexuality.
+This particular module is made up of two questions: one addressing gender identity, and one addressing sexual orientation.
+The question of interest to this paper is that of gender identity, which the BRFSS deliberately asks: “Do you identify as transgender?” If answered yes, the participant then has the options of male-to-female, female-to-male, or gender nonconforming [@Healthcare].
+Here we see the BRFSS having the same initial options as the GSS, but with an additional layer afterwards.
+With this additional set of options, the GSS would likely receive much more detailed data on participant’s identities.
+
+While unfortunately impacted by the COVID-19 pandemic, the 2019 GSS maintains a degree of inclusivity when gathering data on gender identity and sexuality.
+These efforts can be improved upon, we believe, with the application of strategies used by the New Zealand Standard and the BRFSS, which are articulate definitions and a greater variety of options, respectively.
+
+\newpage
 
 ## Appendix: Survey
-Our survey can be found online [here](https://forms.gle/AsAQ8nudjoSHc6sz9).
+Our survey can be found online here: [https://forms.gle/AsAQ8nudjoSHc6sz9 (link)](https://forms.gle/AsAQ8nudjoSHc6sz9).
+
+What follows is a text-based copy of the survey.
+
+---
+
+### Self-identification of sexuality and gender: the terminology of labels and their use
+
+The National Opinion Research Center (NORC) at the University of Chicago has selected you for a supplementary survey, in order to better understand the evolution of the identification and labelling of sexuality (i.e., sexual orientation) and gender in the United States.
+
+By completing this survey, you agree that the NORC may use your responses, stripped of any and all personally-identifying information, in analysis, present and future, in the forum or fora of their choosing.
+The results of this survey will be stored indefinitely in a secure online server, and will be made available to the public via downloadable datasheets.
+This survey is voluntary; there is no penalty for declining to participate, withdrawing from participation before submitting (even if you've already started responded), or not answering any question (not marked as required).
+
+Furthermore, this survey focuses entirely on the verbiage of self-identity in gender and sexuality, not on any physical aspect thereto.
+You will not be asked any personal medical, physiological, or anatomical questions.
+
+#### Introduction and demographics
+
+##### A note on the language of this survey:
+
+For the purposes of this survey, we define a few common terms as follows. Note that these are not end-all definitions, but merely notational for the purpose of communicating questions.
+
+- A "man" refers to any person who identifies as male or masculine.
+- A "woman" refers to any person who identifies as female or masculine.
+- A "non-binary" person refers to any person who identifies as neither male nor female, irrespective of the specific term they may use to this end (agender, gender non-conforming, gender-fluid, etc.)
+- "Cisgender(ed)" refers to a person whose current gender identity corresponds to that which was assigned to them at birth, usually based on physiological traits.
+E.g., a cisgender woman is a woman who was assigned female at birth, and continues to identify as female.
+- "Transgender(ed)" refers to a person whose current gender identity does NOT correspond to their assigned gender at birth.
+E.g., a transgender woman is a woman who was NOT assigned female at birth, but consciously underwent/is undergoing a process of social transition into a woman.
+The gender in question need not be one of the male-female binary; under this definition, non-binary people ARE considered to be transgender.
+- "Attracted to" refers to the experience of experiencing sexual attraction to the relevant party, and is usually assumed to be primary/exclusive unless otherwise stated.
+E.g., someone who is "attracted to men" is primarily/exclusively attracted to people who identify as men.
+Note that we specify sexual attraction; romantic attraction will be the focus of future study.
+- "Bisexual" refers to ANY sexual orientation that affords attraction to more than one gender.
+That is, a bisexual person may e.g., experience attraction to different genders differently or disprefer a particular gender, but they are still capable of experiencing attraction to these genders.
+
+##### Questions
+1. Please describe your gender identity, in whatever terms you use the most frequently.
+2. Has your gender identity ever been different from this identity, for any period of time?
+3. Would you describe yourself as cisgender?
+4. If you answered "No" to the above, do you describe yourself first by your gender identity (man, woman, etc.), or as transgender?
+5. Please enter your preferred NOMINATIVE pronouns, i.e., what would be used to refer to you in the third person. If you use more than one set of pronouns, feel free to include them with slashes.
+6. Please describe your sexual orientation, in whatever terms you use most frequently.
+
+### Opinions and views on the language of identification
+For the following questions, please rate how much you agree with the statement, on a scale of "Strongly agree" to "Strongly disagree".
+
+For questions that ask you about the definitions of words, base your response on how you understand the term and how you use it.
+Disregard, e.g., how other people around you use the term: imagine that you could instantly make everyone use and understand the term the exact same way you do.
+
+##### Questions, contd.
+7. There is sufficient, commonly-understood language to accurately describe my gender and sexuality.
+8. I find the current state of self-identifying language (pronouns, sexuality terms, etc.) confusing.
+9. The word "gay" only refers to gay men, i.e., men who are exclusively sexually attracted to men.
+10. Men who are attracted to men and non-binary people, but not women, are gay.
+11. Straight (i.e., exclusively heterosexual) people can be attracted to non-binary people.
+12. Gay (i.e., exclusively homosexual) people can be attracted to non-binary people.
+13. Women can be gay.
+14. The word "lesbian" only refers to women who are exclusively attracted to women.
+15. Lesbians can be attracted to non-binary people.
+16. A relationship between a transgender person of a binary gender (i.e., male or female), and someone of the opposite binary gender, is straight.
+17. "Bisexual" means attraction to men and women.
+18. "Bisexual" includes attraction to non-binary people and other genders.
+19. A relationship between a bisexual and someone of the same gender is gay.
+20. A relationship between a bisexual and someone of a different gender (e.g., a bisexual man and a woman, regardless of her orientation) is gay.
+21. A relationship between two bisexuals is gay, regardless of the genders of the two involved.
+22. "Bisexual" and "pansexual" mean similar, but ultimately different, things.
+23. Non-binary people are necessarily transgender.
+24. Non-binary people are not cisgender and not transgender, but something different entirely.
+25. Non-binary people can be straight.
+26. Non-binary people can be gay.
+27. A relationship involving a non-binary person is necessarily not a straight relationship.
 
